---
title: "Generalized Estimating Equations (GEE)"
author: "Johnny Hong and Kellie Ottoboni"
categories: [Method]
tags: [generalized estimating equations, linear models]
output: html_document
date: 2017-05-10T21:13:14-05:00
---

Motivation
========================================================

Suppose we observe repeated measurements (responses and/or covariates) on a group of subjects.
We're interested in modeling the expected response for an individual based on these covariates.
Some examples might include

* assigning individuals to one of several controlled diets and measuring their cholesterol over time
* studying the relationship of some variable with earnings over time
* determining the effect of having children on a woman's probability of participation in the labor force

The benefit of having panel data (repeated measurements) like this is that we can control for time-invariant, unobservable differences between individuals.
Having multiple observations per individual allows us to base estimates on the variation *within* individuals.

The easiest way to do answer these questions would be to fit a linear model to the data, where the covariates have an additive effect on the outcome.
If the variables follow something other than a linear relationship (e.g. the response of interest is a probability), a *generalized* linear model (GLM) would be more appropriate.
GLMs have the following form:
$$Y_i = \mu_i + \varepsilon_i, \qquad g(\mu_i) = X_i'\beta$$

where for individual $i$, $Y_i$ is the response, $X_i$ are covariates, $\beta$ is a vector of coefficients, $\varepsilon_i$ is a random error term, and $g$ is a **link function** that maps from the set of possible responses to a linear function of the covariates.

To estimate parameters and do inference with a GLM, we must assume that errors are independent and identically distributed.
With panel data, this clearly isn't the case: observations for each individual are correlated.

As we saw in an earlier presentation, one possible solution is to include [subject-specific random effects](https://rlbarter.github.io/Practical-Statistics/2017/03/03/fixed-mixed-and-random-effects/) in the model fitting.
This method is called a **Generalized Linear Mixed Model (GLMM)**.
GLMMs require some parametric assumptions; if you're like me (Kellie), assuming that everything is Gaussian probably makes you uncomfortable.

Generalized estimating equations (GEE) are a nonparametric way to handle this.
The idea of GEE is to average over all subjects and make a good guess on the *within-subject covariance structure*.
Instead of assuming that data were generated from a certain distribution, uses moment assumptions to iteratively choose the best $\beta$ to describe the relationship between covariates and response.

**Warning:** Notice that I did not specify the objective of the analysis. 
The interpretations of the resultingestimates are different (!) for GLMM and GEE.

Subject-specific versus Population-averaged
========================================================

GEE estimates **population average** effects.
Consider the following two scenarios:

- Scenario 1: You are a doctor. You want to know how much a statin drug will lower your patient’s odds of getting a heart attack.
- Scenario 2: You are a state health official. You want to know how the number of people who die of heart attacks would change if everyone in the at-risk population took the stain drug.

*Source: Allison, P. (2009)*

In the first scenario, we want to know the subject-specific odds.
In the second, we are interested in the prediction for the entire population.
GEE can give us estimates for the second, but not the first.


Nuts and Bolts of GEE
========================================================

GEE estimates **population-averaged** model parameters and their standard errors.
The assumptions for GEE are similar to the assumptions for GLMs:

1. The responses $Y_1, Y_2, \dots, Y_n$ are correlated or clustered
2. There is a linear relationship between the covariates and a transformation of the response, described by the link function $g$.
3. Within-subject covariance has some structure ("working covariance"):
  - independence (observations over time are independent)
  - exchangeable (all observations over time have the same correlation)
  - AR(1) (correlation decreases as a power of how many timepoints apart two observations are)
  - unstructured (correlation between all timepoints may be different)

We need to pick one of these working covariance structures in order to fit the GEE.
As with GLMs, GEE is done using a flavor of [iteratively reweighted least squares](https://en.wikipedia.org/wiki/Iteratively_reweighted_least_squares), plugging in the working covariance matrix as a weight.
The weighted least squares problems we fit are the eponymous **estimating equations**.
If you're familiar with maximum likelihood, you can think of this equation as the [score function](https://en.wikipedia.org/wiki/Score_(statistics)).
This function equals 0 at the optimal choice of $\beta$.

GEE is a **semiparametric method**: while we impose some structure on the data generating process (linearity), we do not fully specify its distribution.
Estimating $\beta$ is purely an exercise in optimization.


What if I am worried that the covariance is misspecified?
========================================================

We have to pick the covariance structure in order to estimate $\beta$, but what if it's not right?

Since the estimating equations are really based on the first moment, $\beta$ will be estimated consistently, even if the working covariance structure is wrong.
However, the standard errors computed from this will be wrong.
To fix this, use GEE with the Huber-White "sandwich estimator" for robustness.
The idea behind the sandwich variance estimator is to use the empirical residuals to approximate the underlying covariance.


Why bother specifying the working covariance to begin with?

  1. Statistical efficiency
  2. Sandwich robustness is a large-sample property


Should we use sandwich all the time?

No, it is problematic if

  1. The number of independent subjects is much smaller than the number of repeated measures
  2. The design is unbalanced -- the number of repeated measures differs across individuals




Example: Pigs
========================================================

**Question:** How does Vitamin E and copper level in the feeds affect the weights of pigs?

Data
  
  - weight of slaughter pigs measured weekly for 12 weeks
  - start weight (i.e. the weight at week 1)
  - cumulated feed intake

Treatments (3x3 factorial design)  
  
  - Vitamin E (dose: 0, 100, 200 mg dl-alpha-tocopheryl acetat/kg feed) 
  - Copper (dose: 0, 35, 175 mg/kg feed)

*Source: Lauridsen, C., Højsgaard, S.,Sørensen, M.T. C. (1999).*


- Implementation in R: `geepack`

```{r, }
library("geepack")
data(dietox)
dietox$Cu <- as.factor(dietox$Cu)
dietox$Evit <- as.factor(dietox$Evit)
mf <- formula(Weight ~ Time + Evit + Cu)
head(dietox)
```

#### Independence Working Covariance

```{r}
geeInd <- geeglm(mf, id=Pig, data=dietox, family=gaussian, corstr="ind")
summary(geeInd)
anova(geeInd)
```

#### Exchangeable Working Covariance


```{r}
geeEx <- geeglm(mf, id=Pig, data=dietox, family=gaussian, corstr="ex")
summary(geeEx)
anova(geeEx)
```

#### AR(1) Working Covariance


```{r}
geeAr1 <- geeglm(mf, id=Pig, data=dietox, family=gaussian, corstr="ar1")
summary(geeAr1)
anova(geeAr1)
```

Advantages
========================================================
- Computationally simple relative to MLE counterparts.
- No distributional assumptions.
- Estimates are consistent even if the correlation structure is misspecified (assuming that the model for the mean response is correct)

Limitations
========================================================
- Likelihood-based methods are not available for usual statistical inference. GEE is a quasi-likelihood method.
- Unclear on how to perform model selection, as GEE is just an estimating procedure. There is no goodness-of-fit measure readily available.
- No subject-specific estimates; if that is the goal of your study, use a different method.

Extensions of GEE
========================================================
- GEE2: second-order extension
  - The GEE version in this presentation is GEE1.
  - Idea: use more complex equations to study the covariance
- Alternating Logistic Regression (ALR) (Carey, Zeger, and Diggle (1993)): model an outcome conditional on another outcome
  - Idea: use log odd ratios instead of correlations to model associations

Take-Home Messages about GEE
========================================================
- ONLY the first the **mean** and the **covariance** matter (quasi-likelihood approach)
- Use a **sandwich estimator** to guard against covariance mispecification
- Model **population-averaged** effects
- Useful when the within-subject dependence is unobserved/unknown
- Still assume subject independence (conditioned on covariates)

References
========================================================

#### GEE
- Liang, K., and S. L. Zeger (1986). Longitudinal data analysis using generalized linear models. Biometrika, 73:13–22.
- Fitzmaurice, G. M., Ware, J.H. and Laird, N. M. (2004). Applied Longitudinal Analysis. Wiley. (Chapter 13)
- Molenberghs, Geert and Verbeke, Geert (2005). Models for Discrete Longitudinal Data. Springer. (Chapter 8)

#### To GEE or not to GEE:
- Hubbard, A.E., Ahern, J., Fleischer, N.L., Van der Laan, M., Lippman, S.A., Jewell, N., Bruckner, T., Satariano, W.A. (2010). To GEE or not to GEE: comparing population average and mixed models for estimating the associations between neighborhood risk factors and health. Epidemiology 21:467–474.

"...We conclude that the estimation-equation approach of population average models provides a more useful approximation of the truth."

#### Subject-specific versus Population-averaged:
Allison, P. D. (2009). Fixed Effects Regression Models (Quantitative Applications in the Social Sciences). SAGE.

#### Blog post:
Dealing with ugly data: Generalized Estimating Equations (GEE) by BOUSTERHOUT:
https://wildlifesnpits.wordpress.com/2014/10/24/dealing-with-ugly-data-generalized-estimating-equations-gee/

#### Dataset:
Lauridsen, C., Højsgaard, S.,Sørensen, M.T. C. (1999). Influence of Dietary Rapeseed Oli, Vitamin E, and Copper on Performance and Antioxidant and Oxidative Status of Pigs. J. Anim. Sci.77:906-916

*Available in the R package geepack*
